HO unification from object language to meta language

### 2024-02-14 main.tex
wip

of course one wants to avoid
\begin{minted}{elpi}
prove P S S'' :- ol-unif P S (forall x\ app "F" x) S', ... assignment "F" S' F, bla F S' ...
\end{minted}

@@

\begin{minted}{elpi}
  kind fo_tm type.
  type fo_app list fo_tm -> fo_tm.
  type fo_lam (tm -> fo_tm) -> fo_tm.
  type fo_uv  nat -> fo_tm.
  type fo_c string -> fo_tm.
\end{minted}

\begin{minted}{elpi}
  kind tm type.
type app list tm -> tm.
type lam (tm -> tm) -> tm.
type uv  nat -> list tm -> tm.
type c string -> tm.
\end{minted}

@@

@@

proprieta desiderate, serve equalho (fa deref)

@@

In our encoding, we explicitly encode the meta-variables with the \elpiIn{ml.uv}
constructor. This is because we prefer to have the full control of the \ml,
including the meta-variables instantiation. This way we are able to concretely
touch the substitution performed by the \ml. In a further section, we show that
there is no difference between our custom \ml language and any other \ml. Of
course, a full control on the unification behind meta-variable assignment ask to
drag the substitution mapping of the \ml and update it each time a variable is
refined.

>>>>

of course one wants to avoid
\begin{minted}{elpi}
prove P S S'' :- ol-unif P S (forall x\ app "F" x) S', ... assmt "F" S' F, bla F S' ...
\end{minted}

@@

\begin{tabular}{c|c}
\begin{minipage}{.5\textwidth}
\begin{minted}{elpi}
kind fo_tm type.
type fo_app list fo_tm -> fo_tm.
type fo_lam (tm -> fo_tm) -> fo_tm.
type fo_uv  nat -> fo_tm.
type fo_c string -> fo_tm.

typeabbrev subst list (option tm).
type equal subst -> tm -> tm -> o.
\end{minted}
\end{minipage}
&
\begin{minipage}{.5\textwidth}
\begin{minted}{elpi}
kind tm type.
type app list tm -> tm.
type lam (tm -> tm) -> tm.
type uv  nat -> list tm -> tm.
type c string -> tm.

kind assmt type.
type abs (tm -> assmt) -> assmt.
type val tm -> assmt.

typeabbrev subst list (option assmt).
type equal subst -> tm -> tm -> o.
\end{minted}
\end{minipage}
\end{tabular}

@@

this should hold in both encoding
%abella
\begin{minted}{elpi}
forall t1 t2 s, unif t1 t2 nil s -> equal s t1 t2
\end{minted}
we have unif only in the HO one, where equal is
weaker (but unif is stronger)

@@

\begin{tabular}{c}
\begin{minipage}{.99\textwidth}
\begin{minted}{elpi}
% deref
equal S (uv N) T1 :- assigned? N S T, equal S T T1.
equal S T1 (uv N) :- assigned? N S T, equal S T1 T.
equal S (app [uv N|Args]) T1 :- assigned? N S F, beta F Args T, equal S T T1.
equal S T1 (app [uv N|Args]) :- assigned? N S F, beta F Args T, equal S T1 T.
% congruence
equal S (app L1) (app L2) :- forall2 (equal S) L1 L2.
equal S (lam F1) (lam F2) :- pi x\ equal S x x => equal S (F1 x) (F2 x).
equal _ (c X) (c X).
equal _ (uv N) (uv N).
% beta
equal S (app [lam X | TL]) T :- beta (lam X) TL T', equal S T' T.
equal S T (app [lam X | TL]) :- beta (lam X) TL T', equal S T T'.
% eta
equal S (lam F) T :- not (T = lam _),
  pi x\ beta T [x] (T' x), equal S (lam F) (lam T').
equal S T (lam F) :- not (T = lam _),
  pi x\ beta T [x] (T' x), equal S (lam T') (lam F).
\end{minted}
\end{minipage}
\\\hline
\begin{minipage}{.99\textwidth}
\begin{minted}{elpi}
% deref
equal S (uv N Args) T1 :- assigned? N S F, beta-abs F Args T, equal S T T1.
equal S T1 (uv N Args) :- assigned? N S F, beta-abs F Args T, equal S T1 T.
% congruence
equal S (app L1) (app L2) :- forall2 (equal S) L1 L2.
equal S (lam F1) (lam F2) :- pi x\ equal S x x => equal S (F1 x) (F2 x).
equal _ (c X) (c X).
equal _ (uv N A1) (uv N A2) :- forall2 (equal S) A1 A2. %fixme in the code
\end{minted}
\end{minipage}

\end{tabular}

equal fo is congruence, deref, beta, eta (for the OL).
equal ho is congruence, deref (for the OL).
unif (fo or ho) should cover pattern fragment, i.e. well behaved unif.

ho, and unif ho, is the LF/ML language/unif. here we describe fo and ho
with a deeply embedded syntax, but the idea is that uv can be mapped to
an actual unif variable.

@@

In our encoding, we explicitly encode the meta-variables with the \elpiIn{ml.uv}
constructor. This is because we prefer to have the full control of the \ml,
including the meta-variables instantiation. This way we are able to concretely
touch the substitution performed by the \ml. In a further section, we show that
there is no difference between our custom \ml language and any other \ml. Of
course, a full control on the unification behind meta-variable assmt ask to
drag the substitution mapping of the \ml and update it each time a variable is
refined.
